---
output:
  pdf_document: default
  html_document: null
  word_document: default
---

# Speedy Analytics
### An analysis of the effects of transmission on car miles per gallon
##### Jim Leach 

********************************************************************************
### Executive Summary

This report addresses two key questions:

1. Is an automatic or manual transmission better for MPG?; and
2. Quantify the difference between automatic and manual transmissions.

To address these two questions, a simple linear regression has been performed 
on the [*mtcars*](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) 
dataset. Exploratory data analysis was performed and a final model selected. 
Diagnostics were performed on the final model and show it to be appropriate for
addressing these questions. 

A simple comparison of the mean miles per gallon between automatic and manual cars
reveals a statistically significant difference ($p$-value less than 0.05).

The model shows that a [TO DO] transmission provides a better (i.e. larger)
miles per gallon value than does automatic. Quantitatively, manual transmission
results in a [TO DO VALUE HERE - include confidence interval!] [TO DO] in
miles per gallon compared to automatic.

********************************************************************************
### Description of approach

#### Data load and exploration:

The set of packages to be used during the analysis (Appendix 2) were loaded and 
the data read in to R. 
```{r loadPack&Dat, echo=FALSE,results='hide',message=FALSE}
library("dplyr", quietly = T)
library("ggplot2", quietly = T)
library("magrittr", quietly = T)
library("GGally", quietly = T)
library("MASS", quietly = T)
library("xtable", quietly = T)

# dir <- "C:/Users/Jleach1/Documents/R/regression_models"
# setwd(dir)
# rm(dir)

data(mtcars)
mtcars$am <- factor(mtcars$am)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$trans <- factor(ifelse(mtcars$am == 0, "Automatic", "Manual"))
```

A quick summary of the differences in miles per gallon accross transmission types
was created and viewed:
```{r summaryDat, echo = F, results = 'asis'}
options(xtable.comment = FALSE)
summ_simple <- mtcars %>%
               mutate(trans = as.factor(ifelse(am==0,"Automatic","Manual"))) %>%
               group_by(trans) %>%
               summarise(Av.MPG = mean(mpg),
                         Stand.Dev = sd(mpg),
                          Number.Cars = n(),
                          Stand.Error = Stand.Dev/sqrt(Number.Cars))
summ_simple[,c(1,2,3,5)] %>% xtable

violin <-
mtcars %>%
  ggplot(aes(x = as.factor(ifelse(am==0,"Automatic","Manual")),
             y = mpg, 
             fill = as.factor(ifelse(am==0,"Automatic","Manual"))))+
  geom_violin()+
  geom_point()+
  theme_minimal()+
  scale_fill_manual(values = c("steelblue","firebrick"))+
  theme(legend.position = "none",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank())+
  xlab("Transmission Type")+
  ylab("Miles per Gallon")

auto <- mtcars[mtcars$am == 0,1]
manual <- mtcars[mtcars$am == 1,1]
t <- t.test(manual, auto)
```

From this summary, we see that the average miles per gallon is `r I(round(summ_simple[1,2],2))` for an `r I(summ_simple[1,1])` car and `r I(round(summ_simple[2,2],2))` for a `r I(summ_simple[2,1])` car. The difference in these means was compared using a $t$-test and was found to lie between
`r I(round(t$conf.int[1],2))` and `r I(round(t$conf.int[2],2))` with a $p$-value of `r I(round(t$p.value,4))`. Therefore the difference in means is `r ifelse(t$p.value <= 0.05, "significant", "insignificant")` and so we `r ifelse(t$p.value <= 0.05,"reject the null hypothesis","fail to reject the null hypothesis")` (i.e. that the difference in means is 0).

A violin plot of the miles per gallon values for each transmission type is shown 
as Figure 1 in Appendix 3.

From the summary and the plot it is seen that that cars with manual transmission
have a greater average mies per gallon than for automatic cars, but a greater 
variance of miles per gallon values. To quantify the strength of this relationship,
and to investigate the effect of other variables on the miles per gallon value, a
model must be created.

#### Model selection:

The [*mtcars*](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
data set contains a number of variables that could have an effect on the miles per
gallon value.

In the context of this report, it is assumed that some assumptions could be made as to
which variables may have a significant effect on a car's miles per gallon (in addition to transmission). These variables are:

* weight (*wt*);
* horsepower (*hp*);
* displacement (*disp*) - a measure of engine size and power;
* number of cyclinders (*cyl*); and
* time to drive 1/4 mile (*qsec*) - another measure of engine power.

The first model that was created simply looked at the relationship between miles per gallon and *all* of these variables to assess which have a large effect on miles per gallon. Rather than 
manually fit multiple models to the data, automated stepwise regression was performed using these variables to find the best combination of them.

```{r initModel, echo=FALSE, results='hide', message=FALSE}
fitAll <- lm(mpg ~ trans + wt + hp + disp + factor(cyl) + qsec ,data = mtcars)
step <- stepAIC(fitAll)
aov <- step$anova
```

This was done using the `stepAIC()` function from the `MASS` package. This gives 
a final model that only includes transmission, weight and *qsec* (time to drive 
1/4 mile) as significant variables to use in the model. Therefore the final model
was built using only these variables and is described by the formula:

$mpg = \beta_{0} + \beta_{1}\times$$am + \beta_{2}\times$$wt + \beta_{3}\times$$qsec$

#### Model Output:

The below table presents the results from the application of this model to the data:

```{r finalModel, echo=FALSE, results = 'asis'}
options(xtable.comment = FALSE)
fitRef <- lm(formula(step), data = mtcars)
intervals <- confint(fitRef)
coefs <- summary(fitRef)$coef %>% as.data.frame %>% cbind(intervals)
coefs <- coefs[,c(1,5,6,2:4)]
coefs %>% xtable

```

The following observations are made:

* Holding the *weight* and *qsec* confounders constant, an increase of `r coefs[2,1] %>% round(2) %>% I()` miles per gallon [`r coefs[2,2] %>% round(2) %>% I()`, `r coefs[2,3] %>% round(2) %>% I()`] is expected for manual cars over automatics.
* This is a statistically significant result with a $p$-value of `r coefs[2,6] %>% round(3)`.
* An increase in *weight* for a given car has a larger absolute effect than transmission, generating a descrease of `r coefs[3,1] %>% round(2) %>% I()`[`r coefs[3,2] %>% round(2) %>% I()`, `r coefs[3,3] %>% round(2) %>% I()`] miles per gallon per unit increase. 
* *qsec* also has a large effect on miles per gallon, with an increase of `r coefs[4,1] %>% round(2) %>% I()` [`r coefs[4,2] %>% round(2) %>% I()`, `r coefs[4,3] %>% round(2) %>% I()`] miles per gallon per 1 second increase in *qsec*.
* The coefficients for *wt* and *qsec* are of an increased statistical significance over transmission, with $p$-values of `r coefs[3,6] %>% round(3)` and `r coefs[4,6] %>% round(3)` respectively.
* The model has an adjusted $R^2$ value of `r (fitRef %>% summary)$r.squared %>% round(3)`, meaning that a large proportion of the variation in the data is being accounted for by the model.

It should be noted that diagnostics of the model fit were performed. 





