---
output:
  pdf_document: default
  html_document: null
  word_document: default
---

# Speedy Analytics
### An analysis of the effects of transmission on car miles per gallon
##### Jim Leach 

********************************************************************************
### Executive Summary

This report addresses two key questions:

1. Is an automatic or manual transmission better for MPG?; and
2. Quantify the difference between automatic and manual transmissions.

To address these two questions, a simple linear regression has been performed 
on the [*mtcars*](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) 
dataset. Exploratory data analysis was performed and a final model selected. 
Diagnostics were performed on the final model and show it to be appropriate for
addressing these questions. 

A simple comparison of the mean miles per gallon between automatic and manual cars
reveals a statistically significant difference ($p$-value less than 0.05). This (simplistically) answers the initial question of which transmission is better for MPG.

Investigating this further to prodce a model that accounts for confounding variables in the data shows that a manual transmission provides a better (i.e. larger) miles per gallon value than does automatic. Quantitatively, manual transmission results in a 2 [0.05, 5.83] miles per gallon increase compared to automatic ($p$-value of 0.047). This further answers the first question, and additionally provides an answer for the second, too.

********************************************************************************
### Description of approach

#### Data load and exploration:

The set of packages to be used during the analysis (Appendix 4) were loaded and 
the data read in to R. 
```{r loadPack&Dat, echo=FALSE,results='hide',message=FALSE}
library("dplyr", quietly = T)
library("ggplot2", quietly = T)
library("gridExtra", quietly = T)
library("magrittr", quietly = T)
library("MASS", quietly = T)
library("xtable", quietly = T)

# dir <- "C:/Users/Jleach1/Documents/R/regression_models"
# setwd(dir)
# rm(dir)

data(mtcars)
mtcars$am <- factor(mtcars$am)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$trans <- factor(ifelse(mtcars$am == 0, "Automatic", "Manual"))
```

A quick summary of the differences in miles per gallon across transmission types
was created and viewed:
```{r summaryDat, echo = F, results = 'asis'}
options(xtable.comment = FALSE)
summ_simple <- mtcars %>%
               mutate(trans = as.factor(ifelse(am==0,"Automatic","Manual"))) %>%
               group_by(trans) %>%
               summarise(Av.MPG = mean(mpg),
                         Stand.Dev = sd(mpg),
                          Number.Cars = n(),
                          Stand.Error = Stand.Dev/sqrt(Number.Cars))
summ_simple[,c(1,2,3,5)] %>% xtable

violin <-
mtcars %>%
  ggplot(aes(x = as.factor(ifelse(am==0,"Automatic","Manual")),
             y = mpg, 
             fill = as.factor(ifelse(am==0,"Automatic","Manual"))))+
  geom_violin(alpha= 0.5)+
  geom_point(alpha= 0.5)+
  theme_minimal()+
  scale_fill_manual(values = c("steelblue","firebrick"))+
  theme(legend.position = "none",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank())+
  xlab("Transmission Type")+
  ylab("Miles per Gallon")

auto <- mtcars[mtcars$am == 0,1]
manual <- mtcars[mtcars$am == 1,1]
t <- t.test(manual, auto)
```

From this summary, we see that the average miles per gallon is `r I(round(summ_simple[1,2],2))` for an `r I(summ_simple[1,1])` car and `r I(round(summ_simple[2,2],2))` for a `r I(summ_simple[2,1])` car. The difference in these means was compared using a $t$-test and was found to lie between `r I(round(t$conf.int[1],2))` and `r I(round(t$conf.int[2],2))` with a $p$-value of `r I(round(t$p.value,4))`. Therefore the difference in means is `r ifelse(t$p.value <= 0.05, "significant", "insignificant")` and so we `r ifelse(t$p.value <= 0.05,"reject the null hypothesis","fail to reject the null hypothesis")` (i.e. that the difference in means is 0).

An exploratory violin plot of the miles per gallon values for each transmission type is shown 
in Appendix 1.

##### Exploratory Analysis Conclusions

From this simple exploratory summary and the plot it is seen that that cars with manual transmission have a greater average miles per gallon than for automatic cars, but a greater variance of miles per gallon values. To quantify the strength of this relationship, and to investigate the effect of other variables on the miles per gallon value, a
model must be created.

#### Model selection:

The [*mtcars*](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) data set contains a number of variables that could have an effect on the miles per gallon value.

In the context of this report, some assumptions could be made as to which variables may have a significant effect on a car's miles per gallon (in addition to transmission). These variables are:

* weight (*wt*);
* horsepower (*hp*);
* displacement (*disp*) - a measure of engine size and power;
* number of cyclinders (*cyl*); and
* time to drive 1/4 mile (*qsec*) - another measure of engine power.

The first model that was created simply looked at the relationship between miles per gallon and *all* of these variables to assess which have a large effect on miles per gallon. Rather than manually fit multiple models to the data, automated stepwise regression was performed using these variables to find the best combination of them.

```{r initModel, echo=FALSE, results='hide', message=FALSE}
fitAll <- lm(mpg ~ trans + wt + hp + disp + factor(cyl) + qsec ,data = mtcars)
step <- stepAIC(fitAll, direction = "both")
aov <- step$anova %>% fortify
```

The output of the model containing all the chosen parameters is found in Appendix 2.

Stepwise regression was done using the `stepAIC()` function from the `MASS` package. This gives a final model that only includes transmission, weight and *qsec* (time to drive 1/4 mile) as significant variables to use in the model. Therefore the final model was built using only these variables and is described by the formula:

$mpg = \beta_{0} + \beta_{1}\times$$am + \beta_{2}\times$$wt + \beta_{3}\times$$qsec$

#### Model Output and Coefficient Interpretation

The below table presents the results from the application of this model to the data:

```{r finalModel, echo=FALSE, results = 'asis'}
options(xtable.comment = FALSE)
fitRef <- lm(formula(step), data = mtcars)
intervals <- confint(fitRef)
coefs <- summary(fitRef)$coef %>% as.data.frame %>% cbind(intervals)
coefs <- coefs[,c(1,5,6,2:4)]
coefs %>% xtable
```

The following observations are made:

* Holding the *weight* and *qsec* confounders constant, an increase of `r coefs[2,1] %>% round(2) %>% I()` miles per gallon [`r coefs[2,2] %>% round(2) %>% I()`, `r coefs[2,3] %>% round(2) %>% I()`] is expected for manual cars over automatics.
* This is a statistically significant result with a $p$-value of `r coefs[2,6] %>% round(3)`.
* An increase in *weight* for a given car has a larger absolute effect than transmission, generating a decrease of `r coefs[3,1] %>% round(2) %>% I()`[`r coefs[3,2] %>% round(2) %>% I()`, `r coefs[3,3] %>% round(2) %>% I()`] miles per gallon per unit increase. 
* *qsec* also has a large effect on miles per gallon, with an increase of `r coefs[4,1] %>% round(2) %>% I()` [`r coefs[4,2] %>% round(2) %>% I()`, `r coefs[4,3] %>% round(2) %>% I()`] miles per gallon per 1 second increase in *qsec*.
* The coefficients for *wt* and *qsec* are of an increased statistical significance over transmission, with $p$-values of `r coefs[3,6] %>% round(3)` and `r coefs[4,6] %>% round(3)` respectively.
* The model has an adjusted $R^2$ value of `r (fitRef %>% summary)$r.squared %>% round(3)`. Given that the model has only three variables in it, this is indicative of a good model fit.

Additional diagnostics of the model fit were performed (Appendix 3) and show that, broadly, the model fit is good and is appropriate for the exercise at hand.

********************************************************************************

### Appendices
#### Appendix 1 - Violin plot of miles per gallon by transmission type

```{r appendix1, echo=FALSE, fig.align='center'}
violin
```


#### Appendix 2 - Model parameters from unrefined model with all chosen variables


```{r appendix2, echo=FALSE, results = 'asis', fig.align='center'}
options(xtable.comment = FALSE)
(fitAll %>% summary)$coef %>% cbind(fitAll %>% confint) %>% xtable
```

#### Appendix 3 - Diagnostic Plots
```{r appendix3, echo=FALSE}
# Create a data frame for ggplots
fitRefRes <- fortify(fitRef)

# residuals vs. fitted values
  residFit <- ggplot(fitRefRes, aes(.fitted, .resid))+
    geom_point(colour = "steelblue")+
    geom_smooth(method = "loess", se=FALSE, colour = "firebrick")+
    geom_hline(linetype=2, size=.2)+
    scale_x_continuous("Fitted Values")+
    scale_y_continuous("Residual")+
    ggtitle("Residuals vs Fitted")+
    theme_minimal()
  
# normal Q-Q plot
  a <- quantile(fitRefRes$.stdresid, c(0.25, 0.75))
  b <- qnorm(c(0.25, 0.75))
  slope <- diff(a)/diff(b)
  int <- a[1] - slope * b[1]
  qq <- ggplot(fitRefRes, aes(sample=.resid)) +
    stat_qq(colour = "steelblue") +
    geom_abline(slope=slope, intercept=int, colour = "firebrick") +
    scale_x_continuous("Theoretical Quantiles") +
    scale_y_continuous("Standardized Residuals") +
    theme_minimal()+
    ggtitle("Normal Q-Q")
   
# Scale -location (stanardised resid vs. fitted)
  standFit <- ggplot(fitRefRes, aes(.fitted, sqrt(abs(.stdresid)))) +
      geom_point(colour = "steelblue") +
      geom_smooth(se=FALSE, method = "loess", colour = "firebrick") +
      scale_x_continuous("Fitted Values") +
      scale_y_continuous("Root of Standardized Residuals") +
      theme_minimal()+
      ggtitle("Scale-Location")
  
# residuals vs. leverage
  residLev <- ggplot(fitRefRes, aes(.hat, .stdresid)) +
      geom_point(colour = "steelblue") +
      geom_smooth(se=FALSE, colour = "firebrick", method = "loess") +
      geom_hline(linetype=2, size=.2) +
      scale_x_continuous("Leverage") +
      scale_y_continuous("Standardized Residuals")+
      ggtitle("Residuals vs. Leverage")+
      theme_minimal()

plots <- list(residFit, qq, standFit, residLev)
do.call("grid.arrange",c(plots))
```

#### Appendix 4 - Packages and contact information

* [dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)
* [ggplot2](http://ggplot2.org/)
* [gridExtra](http://cran.r-project.org/web/packages/gridExtra/gridExtra.pdf)
* [magrittr](https://github.com/smbache/magrittr)
* [MASS](http://cran.r-project.org/web/packages/MASS/MASS.pdf)
* [xtable](http://cran.r-project.org/web/packages/xtable/xtable.pdf)

The author of this report can be contacted on [twitter](https://twitter.com/leach_jim)














